Something that is not PCADSC

### 2017-02-25 article/article_v0.01.tex
Ny beskrivelse af PCA

\subsection{Principal component analysis - a very brief introduction}
Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$. This is in itself not very impressive nor useful. But PCA does not result in just any basis of the data matrix. The basis chosen using PCA has the very favorable property that each vector is chosen such that they maximize the accumulated explained variance. These vectors are usually denoted \textit{principal component scores} or simply \textit{principal components} and they are constructed as linear combinations of the variables in the dataset (or, equivalently, as projections of the data matrix). The coefficients of each variable in these projections are denoted \textit{loadings}. Let $S \in \RR^{d \times d}$ denote the empirical covariance matrix and let $X \in \RR^{n \times d}$ denote our observed data matrix. $d$ is the number of variables while $n$ is the number of observations. The deconstruction procedure then goes as follows:
\begin{enumerate}
\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible. More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$. Let $\lambda_1 = \V(\eta_1^T X^T)$. 
\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
\vdots 
\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
\end{enumerate}
The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively. Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions. We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.

 In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.

>>>>

\subsection{Principal component analysis - a very brief introduction}

{\color{red}

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of the subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different. If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{i=q+1}^d \lambda_i$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called the \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called the \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon: u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
\end{align*}

It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and associated variation components $\lambda_j$ describe the structure of the dataset simultaneously for all approximating dimension $q$. This implies that the loadings and variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension $q$.
}

%Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$. This is in itself not very impressive nor useful. But PCA does not result in just any basis of the data matrix. The basis chosen using PCA has the very favorable property that each vector is chosen such that they maximize the accumulated explained variance. These vectors are usually denoted \textit{principal component scores} or simply \textit{principal components} and they are constructed as linear combinations of the variables in the dataset (or, equivalently, as projections of the data matrix). The coefficients of each variable in these projections are denoted \textit{loadings}. Let $S \in \RR^{d \times d}$ denote the empirical covariance matrix and let $X \in \RR^{n \times d}$ denote our observed data matrix. $d$ is the number of variables while $n$ is the number of observations. The deconstruction procedure then goes as follows:
%\begin{enumerate}
%\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible. More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$. Let $\lambda_1 = \V(\eta_1^T X^T)$. 
%\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
%\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
%\vdots 
%\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{\eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
%\end{enumerate}
%The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively. Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions. We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.
%
% In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.
